GPU-friendly Stroke Expansion

### 2024-06-13 paper.tex
Fix citation formatting and typos
This commit applies nearly almost all of the typos and suggested wording changes of reviewer 1.

Correct stroke rendering also relies on accurate approximation of the parallel curves of the input paths. Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula\citenum{Farouki1990}. Thus, all practical stroke rendering techniques employ an approximation, with error tolerance as a tunable parameter, typically representing the maximum allowable Fréchet distance between the true curve and its approximation. A typical value is 0.25 device pixel, as that is close to the threshold of visibility. A correct algorithm will produce an approximate curve within the error tolerance, but an efficient algorithm will not subdivide significantly more than is necessary to meet the error bound. Some specifications for vector graphics, including PDF\citenum{PDF2008}, specify the error tolerance explicitly.

@@

Previous work on stroke rendering does not fully solve the problem of fast and correct stroke rendering. Most existing implementations have correctness problems, in particular failing to draw the evolutes and inner joins as required for strong correctness; \citet{Nehab2020} provides a detailed survey. In addition, most implementations are not suitable for running on GPU, requiring sequential execution to produce watertight outlines. One GPU implementation is polar stroking (\cite{Kilgard2020}), a local technique, but among other limitations, it does not bound the Fréchet distance error (rather, it bounds the angle error).

@@

The core problem in stroke expansion is approximating the desired curve by segments of some other curve, usually a simpler one. These segments must be within an error tolerance of the source curve, and ideally close to a minimal number of them. We consider a number of source to target pairs, most importantly cubic Béziers to Euler spirals, and Euler spiral parallel curves to either lines or arcs.

@@

By far the most efficient approach is an \emph{invertible} error metric. In this approach, the error metric has an analytic inverse, or at least a good numerical approximation. Because the metric is invertible, it can predict the number of subdivisions needed, as well as the parameter value for each subdivision. If the error metric is accurate, then approximation is near-optimal. One example of an invertible error metric is angle step, used in polar stroking (\citet{Kilgard2020}); the number of subdivisions is the total angle subsumed by the curve divided by the angle step size, and the parameter value for each subdivision is the result of solving for a tangent direction. Another widely used invertible error metric is Wang's formula (\citet{Goldman2003}, Section 5.6.3), which gives a bound on the flattening error based on the second derivative of the curve. This metric is conservative but works well in practice; among other applications, it is used in Skia for path flattening. One limitation of Wang's method is that it only applies to the flattening of filled outlines; when applied naively to generation of parallel curves, it can undershoot substantially, especially near cusps.

@@

\begin{figure}
    \includegraphics[scale=0.3]{es_flatten}
    \caption{Flattening of an Euler spiral using points spaced by power-law}
    \Description{An Euler spiral along with its flattened representation. The flattened curve has an error}
    \label{fig:es_flatten}
\end{figure}

@@

The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago[\citenum{Wieleitner1907}], and has a straightforward closed-form Cesàro representation, curvature as a function of arc length.

@@

For a convex curve, the optimal flattening is tractable to compute, though not necessarily very fast. When a curve is monotonic, error is also monotonic with respect to both subdivision points. Thus, any standard root finding technique can find a subdivision with a given error (bisection is the best known, but the ITP method \citet{Oliveira2020} is better). A similar outer loop with a root finding technique can find the error which minimizes the global error for the entire flattening, which, due to monotonicity, is that for which the error of the final segment equals that of the previous segments. Subdividing at the inflection point adds at most one additional point; it is possible to optimize further, but we do not consider that (though the near-optimal flattening we will later present is not subject to this constraint).

@@

The subdivision density for the parallel curve of an Euler spiral, normalized so that its inflection point is at -1 and the cusp of the parallel curve is at 1, is simply $1 - \sqrt{|1-s^2|}$. This function is plotted in Figure~\ref{fig:subdiv_density}.

@@

For an Euler spiral segment, $\kappa'(s)$ is constant and thus this error metric becomes nearly trivial. With n subdivisions, the estimated distance is simply $\frac{s^3\kappa'}{120n^3}$. Solving for $n$, we get $n = s\sqrt[3]{\frac{|\kappa'|}{120d}}$ subdivisions, and those are divided evenly by arc length, as the subdivision density is constant across the curve, just as is the case for flattening arcs to lines.

@@

One consequence of this formula is that, since $h$ appears under absolute value, the same arc approximation can be used for both sides of a stroke.

@@

Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type, more tractable for evaluating parallel curve. Computing parallel curves directly on cubic Bézier curve segments is not very tractable. In particular, the widely cited Tiller-Hanson algorithm[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.

@@

The entire recursion stack for adaptive subdivision can be represented in two words: dt and scaled\_t0. The range is dt * scaled\_t0 .. dt * (scaled\_t0 + 1). Initial values of 1 and 0 represent the range (0, 1). Pushing the stack, subdividing the range in half, is represented by halving dt and doubling scaled\_t0, which leaves the start of the range invariant but its size halved. After accepting an approximation, the next range is determined by incrementing scaled\_t0, then repeatedly popping the stack by doubling dt and halving scaled\_t0, as long as the latter is odd. This can in fact be achieved without iteration by using the ``count trailing zeros'' intrinsic.

@@

Our approach is to leverage the conversion to Euler spiral segments, which have finite curvature. The geometric Hermite interpolation depends on accurate tangents at the endpoints, which in turn are derived from derivatives. The tangents are not well defined when the derivative is zero, and are not numerically stable when near-zero. Thus, the algorithm has one major mechanism to deal with numerical robustness in these cases: when sampling the cubic Bézier, if the derivative is near zero (as determined by testing against an epsilon threshold), then the derivative is sampled at a slightly perturbed parameter value.

@@

An SVG path consists of a sequence of instructions (or "verbs"). The \emph{lineto} and \emph{curveto} instructions denote an individual \emph{path segment} consisting of either a straight line or a cubic Bézier. A given subsequence of path segments can be bounded by a \emph{moveto} and/or \emph{closepath} instruction which mark the beginning and end of a \emph{subpath}. A subpath must form a closed contour if painted as a fill. When painted as a stroke, each subpath must begin and end with a cap according to the desired cap style. In addition, each path segment in a stroked subpath must be connected to adjacent path segments of the same subpath with a join. For any given path, the cap and join styles do not vary across subpaths or individual path segments.

@@

We authored two additional test scenes in order stress the GPU with a higher workload. The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus. We used two variants of this scene with \emph{butt} and \emph{round} cap styles. For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths. We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively. Table~\ref{table:bump-counts} shows the payload sizes of our largest test scenes.

@@

The test scenes were rendered to a GPU texture with dimensions 2088x1600. The contents of each scene were uniformly scaled up (by encoding a transform) to fit the size of the texture, in order to increase the required number of output primitives. As with the CPU timings, we used an error tolerance of 0.25 in all cases. We recorded the execution times across 3,000 GPU submissions for each scene.

\begin{figure}
    \includegraphics[scale=0.69]{nehab_gpu_timings}
    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the \citet{Nehab2020} \emph{timings} data set. The timings are shown in milliseconds. The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs. Mali-G78 (the lowest end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in ~3ms. The compute shader executes at least an order of magnitude faster on the other GPUs.}
    \Description{A stacked bar chart of the GPU timings for stroke expansion from \citet{Nehab2020} timings SVGs}
    \label{fig:nehab-gpu-timings}
\end{figure}

@@

Figure~\ref{fig:tag-monoid-gpu-timings} shows the run time of the tag monoid parallel prefix sums with increasing input size. The prefix sums scale better at large input sizes compared to our stroke expansion kernel. This can be attributed to a lack of expensive computations and high control-flow uniformity in the tag monoid kernel. The CPU encoding times (prior to the GPU prefix sums) for some of the large scenes is shown in Table~\ref{table:bump-estimate-cost}. Dashing on the GPU was out of scope, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequantially encoded as individual strokes. Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.

@@

We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment. We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}). For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth. We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.

@@

\begin{table}
    \caption{The estimated and actual \emph{line soup} buffer sizes for scenes rendered at 2088x1600 resolution. Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth. In the worst case, our buffer size estimator overshoots by a factor ~4.}
    \begin{tabular}{|l|c|c|}
    \hline
        \multicolumn{1}{|c|}{\textbf{Test}} & Actual Size & Estimated Size \\
    \hline
        roads.svg  & 958.78 KB & 2.45 MB    \\
        waves.svg  & 9.37 MB   & 10.44 MB   \\
        mmark-70k  & 23.81 MB  & 100.79 MB  \\
        mmark-120k & 40.86 MB  & 173.56 MB  \\
    \hline
    \end{tabular}
    \label{table:bump-estimate-mem}
\end{table}

@@

The Euler spiral is represented by $\kappa(s) = k_0 + k_1s$, where $s$ ranges from 0 to 1, i.e. the arc length is unit normalized. This segment is scaled, rotated, and translated into place so that its endpoints match the desired locations. Immediately, $k_0$ can be determined as $\theta_0 + \theta_1$. Therefore, the remaining parameters to compute are $k_1$ and the ratio of chord length to arc length. The latter \emph{could} be computed from $k_0$ and $k_1$ by numerical integration, but it is more efficient to obtain it at the same time as $k_1$. Setting $k = \theta_0 + \theta_1$ and $\Delta = \theta_1 - \theta_0$, we have:

>>>>

Correct stroke rendering also relies on accurate approximation of the parallel curves of the input paths. Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula[\citenum{Farouki1990}]. Thus, all practical stroke rendering techniques employ an approximation, with error tolerance as a tunable parameter, typically representing the maximum allowable Fréchet distance between the true curve and its approximation. A typical value is 0.25 device pixel, as that is close to the threshold of visibility. A correct algorithm will produce an approximate curve within the error tolerance, but an efficient algorithm will not subdivide significantly more than is necessary to meet the error bound. Some specifications for vector graphics, including PDF[\citenum{PDF2008}], specify the error tolerance explicitly.

@@

Previous work on stroke rendering does not fully solve the problem of fast and correct stroke rendering. Most existing implementations have correctness problems, in particular failing to draw the evolutes and inner joins as required for strong correctness; \citet{Nehab2020} provides a detailed survey. In addition, most implementations are not suitable for running on GPU, requiring sequential execution to produce watertight outlines. One GPU implementation is polar stroking (\citet{Kilgard2020}), a local technique, but among other limitations, it does not bound the Fréchet distance error (rather, it bounds the angle error).

@@

The core problem in stroke expansion is approximating the desired curve by segments of some other curve, usually a simpler one. These segments must be within an error tolerance of the source curve, and ideally close to a minimal number of them. We consider a number of source-to-target pairs, most importantly cubic Béziers to Euler spirals, and Euler spiral parallel curves to either lines or arcs.

@@

By far the most efficient approach is an \emph{invertible} error metric. In this approach, the error metric has an analytic inverse, or at least a good numerical approximation. Because the metric is invertible, it can predict the number of subdivisions needed, as well as the parameter value for each subdivision. If the error metric is accurate, then approximation is near-optimal. One example of an invertible error metric is angle step, used in polar stroking (\citet{Kilgard2020}); the number of subdivisions is the total angle subsumed by the curve divided by the angle step size, and the parameter value for each subdivision is the result of solving for a tangent direction. Another widely used invertible error metric is Wang's formula (\citet{Goldman2003}, Section 5.6.3), which gives a bound on the flattening error based on the second derivative of the curve. This metric is conservative but works well in practice; among other applications, it is used in Skia for path flattening. The chief limitation of Wang's method is that it computes a subdivision bound for a Bézier curve without accounting for the displacement of the parallel curve due to stroking. When applied naively to generation of parallel curves, it can undershoot substantially, especially near cusps.

@@

\begin{figure}
    \includegraphics[scale=0.3]{es_flatten}
    \caption{The flattened approximation to an Euler spiral using line segments is achieved by choosing subdivision points using a simple power law formula with an exponent of 2/3.}
    \Description{An Euler spiral along with its flattened representation. The flattened curve has a consistent error, and the subdivision points are spaced according to a}
    \label{fig:es_flatten}
\end{figure}

@@

The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length, of the form $\kappa(s) = c_0{s - s_0}^{-1/2} + c_1$.

@@

For a convex curve, the optimal flattening is tractable to compute, though not necessarily very fast. When a curve is monotonic, error is also monotonic with respect to both subdivision points. Thus, any standard root finding technique can find a subdivision with a given error (bisection is the most popular, but the ITP method[\citenum{Oliveira2020}] is better). A similar outer loop with a root finding technique can find the error which minimizes the global error for the entire flattening, which, due to monotonicity, is that for which the error of the final segment equals that of the previous segments. Subdividing at the inflection point adds at most one additional point; it is possible to optimize further, but we do not consider that (though the near-optimal flattening we will later present is not subject to this constraint).

@@

The subdivision density for the parallel curve of an Euler spiral, normalized so that its inflection point is at $-1$ and the cusp of the parallel curve is at 1, is simply $1 - \sqrt{|1-s^2|}$. This function is plotted in Figure~\ref{fig:subdiv_density}.

@@

For an Euler spiral segment, $\kappa'(s)$ is constant and thus this error metric becomes nearly trivial. With $n$ subdivisions, the estimated distance is simply $\frac{s^3\kappa'}{120n^3}$. Solving for $n$, we get $n = s\sqrt[3]{\frac{|\kappa'|}{120d}}$ subdivisions, and those are divided evenly by arc length, as the subdivision density is constant across the curve, just as is the case for flattening arcs to lines.

@@

One consequence of this formula is that, since the error is in terms of the absolute value of $h$, independent of sign, the same arc approximation can be used for both sides of a stroke.

@@

Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type that is more tractable for evaluating parallel curve. Computing parallel curves directly on cubic Bézier curve segments is not very tractable. In particular, the widely cited Tiller-Hanson algorithm[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.

@@

The entire recursion stack for adaptive subdivision can be represented in two words: $dt$ and $\mbox{scaled}\_t0$. The range is $dt \cdot \mbox{scaled}\_t0 .. dt \cdot (\mbox{scaled}\_t0 + 1)$. Initial values of $1$ and $0$ represent the range $(0, 1)$. Pushing the stack, subdividing the range in half, is represented by halving $dt$ and doubling $\mbox{scaled}\_t0$, which leaves the start of the range invariant but its size halved. After accepting an approximation, the next range is determined by incrementing $\mbox{scaled}\_t0$, then repeatedly popping the stack by doubling $dt$ and halving $\mbox{scaled}\_t0$, as long as the latter is odd. This can in fact be achieved without iteration by using the ``count trailing zeros'' intrinsic.

@@

Our approach leverages the conversion to Euler spiral segments, which have finite curvature. The geometric Hermite interpolation depends on accurate tangents at the endpoints, which in turn are derived from derivatives. The tangents are not well defined when the derivative is zero, and are not numerically stable when near-zero. Thus, the algorithm has one major mechanism to deal with numerical robustness in these cases: when sampling the cubic Bézier, if the derivative is near zero (as determined by testing against an epsilon threshold), then the derivative is sampled at a slightly perturbed parameter value.

@@

An SVG path consists of a sequence of instructions (or "verbs"). The \emph{lineto} and \emph{curveto} instructions denote an individual \emph{path segment} consisting of either a straight line or a cubic Bézier. A given subsequence of path segments can be bounded by a \emph{moveto} and/or \emph{closepath} instruction that marks the beginning and end of a \emph{subpath}. A subpath must form a closed contour if painted as a fill. When painted as a stroke, each subpath must begin and end with a cap according to the desired cap style. In addition, each path segment in a stroked subpath must be connected to adjacent path segments of the same subpath with a join. For any given path, the cap and join styles do not vary across subpaths or individual path segments.

@@

We authored two additional test scenes in order to stress the GPU with a higher workload. The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus. We used two variants of this scene with \emph{butt} and \emph{round} cap styles. For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths. We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively. Table~\ref{table:bump-counts} shows the payload sizes of our largest test scenes.

@@

The test scenes were rendered to a GPU texture with dimensions 2088x1600. The contents of each scene were uniformly scaled up (by encoding a transform) to fit the size of the texture, in order to increase the required number of output primitives. As with the CPU timings, we used an error tolerance of 0.25 pixels in all cases. We recorded the execution times across 3,000 GPU submissions for each scene.

\begin{figure}
    \includegraphics[scale=0.69]{nehab_gpu_timings}
    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the \citet{Nehab2020} \emph{timings} data set. The timings are shown in milliseconds. The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs. Mali-G78 (the lowest-end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in \~3~ms. The compute shader executes at least an order of magnitude faster on the other GPUs.}
    \Description{A stacked bar chart of the GPU timings for stroke expansion from \citet{Nehab2020} timings SVGs}
    \label{fig:nehab-gpu-timings}
\end{figure}

@@

Figure~\ref{fig:tag-monoid-gpu-timings} shows the run time of the tag monoid parallel prefix sums with increasing input size. The prefix sums scale better at large input sizes compared to our stroke expansion kernel. This can be attributed to a lack of expensive computations and high control-flow uniformity in the tag monoid kernel. The CPU encoding times (prior to the GPU prefix sums) for some of the large scenes is shown in Table~\ref{table:bump-estimate-cost}. Our implementation still applies dashing on the CPU. It is conceptually fairly straightforwad to implement on GPU; the core algorithm is prefix sum applied to segment length, and the Euler spiral representation is particularly well suited. However, there are complex implementatoin tradeoffs, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequentially encoded as individual strokes. Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.

@@

We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment. We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}). For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth. We observed that enabling estimation increases our CPU pre-processing time by 2--3x on the largest scenes but the overall impact is negligible when the scene size is modest.

@@

\begin{table}
    \caption{The estimated and actual \emph{line soup} buffer sizes for scenes rendered at 2088x1600 resolution. Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth. In the worst case, our buffer size estimator overshoots by a factor of \~4.}
    \begin{tabular}{|l|c|c|}
    \hline
        \multicolumn{1}{|c|}{\textbf{Test}} & Actual Size & Estimated Size \\
    \hline
        roads.svg  & 958.78 KB & 2.45 MB    \\
        waves.svg  & 9.37 MB   & 10.44 MB   \\
        mmark-70k  & 23.81 MB  & 100.79 MB  \\
        mmark-120k & 40.86 MB  & 173.56 MB  \\
    \hline
    \end{tabular}
    \label{table:bump-estimate-mem}
\end{table}

@@

The Euler spiral is represented by $\kappa(s) = k_0 + k_1s$, where $s$ ranges from 0 to 1, i.e., the arc length is unit normalized. This segment is scaled, rotated, and translated into place so that its endpoints match the desired locations. Immediately, $k_0$ can be determined as $\theta_0 + \theta_1$. Therefore, the remaining parameters to compute are $k_1$ and the ratio of chord length to arc length. The latter \emph{could} be computed from $k_0$ and $k_1$ by numerical integration, but it is more efficient to obtain it at the same time as $k_1$. Setting $k = \theta_0 + \theta_1$ and $\Delta = \theta_1 - \theta_0$, we have:
